A Framework for Discrete Change

### 2008-11-28 papers/individual/individual.tex
Updated individual paper with Brenda's comments
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@245 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

%&LaTeX
\documentclass[a4paper,12pt,man,english]{apa} % nobf, doc, apa

\usepackage[english]{babel} % ik krijg anders foute citaties...in Zweeds formaat???

@@

\twoaffiliations{Department of Psychology, University of Amsterdam}{Department of Cognitive, Perceptual and Brain Sciences, University College London}

@@

% balance scale rules
\citet{Jansen2001} applied the catastrophe model to
development of reasoning on the balance scale task (Siegler, 1981).
In the balance scale task participants have to judge which side of a
balance goes down when the number of weights and their distances to
the fulcrum are varied over trials.  Younger children tend to ignore
the distance dimension in this task, and instead focus solely on the
number of weights on each side of the fulcrum.  This strategy for
solving balance scale items is called Rule 1 \cite{Siegler1981}.  Older
children include the distance dimension in determining their response
to balance scale problems; however, they only do so when the weight
dimension does not differ between the sides of the balance, i.e., when
the number of weights is equal on both sides of the balance scale.
This strategy is called Rule 2 \cite{Siegler1981}.

@@

% conditioning and addiction research
Also in animal learning and conditioning, evidence is found for sudden
changes in response behavior \cite{Gallistel2004}.  In particular, in
their study, evidence was found for sudden onset of learning: at the
start of the learning experiment, pigeons did not learn anything and
performance was stable; after a number of trials, learning kicked in
and there were large increases in performance.
\citeauthor{Gallistel2004} focused on modeling the distribution of
onset times: that is, the trials at which learning suddenly takes off.
A similar interest in process onset times is found in addiction
research.  For example, \citet{Sher2004} study the age at which
children start using alcohol and how this related to eventual outcomes
in terms of addiction.

% discrimination and categorization learning
Sudden transitions in learning are also observed in simple
discrimination learning paradigms in which participants learn to
discriminate a number of stimuli based on a single dimension such as
form or color.  This kind of learning is referred to as all-or-none
learning or concept identification learning.  \citet{Raijmakers2001}
found evidence for different strategies applied by children when faced
with such a learning task.  \citet{Schmittmann2006} reanalyzed their
data using hidden Markov models to show that both strategies are
characterized by sudden transitions in the learning process.

% latent Markov versus time series analyses/outline
In above mentioned applications, the data consist mostly of a few
repeated measurements administered to large groups of participants.
The focus in the current chapter is rather on data that consist of
many repeated measurements, or time series, observed in only a few
participants or even just a single participant.  For example,
\citet{Visser2009} analyzed data from a single participant in an
experimental task that manipulates the trade-off between speed and
accuracy.  The data consisted of three time series with each around
150 repeated measurements of both reaction time and accuracy.  Below
we provide examples of analyzing time series from single participants
from two experiments; one from the Iowa Gambling Task and one from the
Weather Prediction Task.  The interest in these tasks is to show that
participants develop different strategies over time in responding to
the stimuli and that the transition from one strategy to the next is a
discrete event.  Before providing these illustrations, below we give a
formalization of dependent mixture models and a brief overview of the
\pkg{depmixS4} package that was developed to specify and fit such models.

@@

% history of Markov models/LMM applications
Markov models have been used extensively in the social sciences; for
example, in analyzing language learning \cite{Miller1952,Miller1963} and
in the analysis of paired associate learning \cite{Wickens1982}.  In
these models, the focus is on survey type data: a few repeated
measurements taken in a large sample.  \citeA{Langeheine1990} discuss
latent Markov models and their use in sociology and political science
(see also McCutcheon, 1987).  Latent transition models have been used
in studying development of math skills \citep{Collins1992} and in
medical applications \citep{Reboussin1998}; \citet{Kaplan2008}
provides an overview of such models, that are called stage-sequential
models in the developmental psychology literature.

@@

% hidden Markov models
Hidden Markov models (HMM) tend to be used in the analysis of long
univariate (individual) timeseries.  For example, HMMs are the model
of choice in speech recognition applications \cite{Rabiner1989}.  In
biology, HMMs are used to analyze DNA sequences \cite{Krogh1998} and
in econometric science, to analyze changes in stock market prices and
commodities \cite{Kim1994}.

% depmix model
The dependent mixture model that we propose here spans the range from
latent Markov models for few repeated measurements with many
participants to hidden Markov models for individual time series.  In
addition, the dependent mixture model includes multivariate responses.
The dependent mixture model consists of the following elements:
\begin{enumerate}
	\item $S$ is a collection of discrete states
	\item $S_{t} = \mat{A}S_{t-1}+\xi_{t}, \mat{A}$, a transition matrix
	\item $O_{t} = \mat{B}(S_{t}) + \zeta_{t}, \mat{B}$,  an observation density
\end{enumerate}
Here $\xi_{t}$ and $\zeta_{t}$ are independent error processes 
\cite{Elliott1995}.

% state space
The state space, which is a set of discrete states, captures the
different states that the learning or developmental process under
consideration can be in.  In the balance scale example mentioned
above, children are applying one of two possible strategies in
responding to the items.  The states are characterized by their
corresponding observation densities.  Using for example Rule 1 in the
balance scale task leads to correct answers on some items and
incorrect answers on others.  A different strategy may lead to correct
answers on some items and to guessing behavior on other items.  In
analyzing categorization learning data, in which participants learn to
categorize a set of objects, a typical initial state is that
participants are guessing because at the start of the task they have
no knowledge of which features are important in categorization.
Hence, the states in the state space represent knowledge states of the
participants, such that different knowledge states lead to different
observed behavior or responses.

% transitions/matrix
The transition matrix $\mat{A}$ describes the transitions between
states over repeated measures or trials.  This matrix summarizes the
probabilities of transitioning from one state to another which
represents learning or development.  The transition model contains the
Markov assumption:
$$Pr(S_{t}|S_{t-1}, \ldots, S_{1}) = Pr(S_{t}|S_{t-1}),$$
which means that the current state (at time $t$) only depends on the
previous state $S_{t-1}$, and not on earlier states.

% observation densities
The observation densities $\mat{B}$ form the measurement part of the
model; these describe the distributions of the observations
conditional on the current state. Hence, these distributions
characterize the state, and in our examples, these characterize the
strategy that a participant is using at a given measurement occasion.

% log-likelihood
The log-likelihood of DMMs is usually computed by the
so-called forward-backward algorithm \citep{Baum1966,Rabiner1989}, or
rather by the forward part of this algorithm.  \citet{Lystig2002}
changed the forward algorithm in such a way as to allow computing the
gradients of the log-likelihood at the same time.  They start by
rewriting the likelihood as follows (for ease of exposition the
dependence on the model parameters is dropped here):
\begin{equation}
	L_{T} = Pr(\vc{O}_{1}, \ldots, \vc{O}_{T}) = \prod_{t=1}^{T}
Pr(\vc{O}_{t}|\vc{O}_{1},
	\ldots, \vc{O}_{t-1}),
	\label{condLike}
\end{equation}
where $Pr(\vc{O}_{1}|\vc{O}_{0}):=Pr(\vc{O}_{1})$. Note that for a
simple, i.e.\ observed, Markov chain these probabilities reduce to
$Pr(\vc{O}_{t}|\vc{O}_{1},\ldots,
\vc{O}_{t-1})=Pr(\vc{O}_{t}|\vc{O}_{t-1})$.
The log-likelihood can now be expressed as:
\begin{equation}
	l_{T} = \sum_{t=1}^{T} \log[Pr(\vc{O}_{t}|\vc{O}_{1}, \ldots,
\vc{O}_{t-1})].
	\label{eq:condLogl}
\end{equation}

@@

	\item to be able to fit transition models with covariates, i.e.,
	to have time-dependent transition matrices.

@@

% response models
Response models in \pkg{depmixS4} interface with the \code{glm} 
functions available in R and with the \pkg{nnet}-package for the 
multinomial logistic models. Normally distributed data can hence be 
modelled with direct effects included as well. For example, normal 
data are modelled as:
\begin{equation}
	O_{t}|S_{t} = \mu + \mathbf{\beta}^{`}\vc{x} + \epsilon_{t},
\end{equation}
where $\mu$ the (state-dependent) mean; $\vc{x}$ is a vector of
covariates or predictors; $\mathbf{\beta}$ is the vector of
(state-dependent) regression coefficients; $\epsilon_{t}$ is a
normally distributed error with (state-dependent) standard deviation
$\sigma$.

@@

@@

% \subsection{Toy data: parameter retrieval} ????

@@

% what is the IGT?
The Iowa gambling task (IGT) is an experimental paradigm designed to
mimic real-life decision-making situations \cite{Bechara1994}, in the
way that it factors uncertainty, reward and punishment
\cite{Dunn2006}.  The task requires the selection of cards from four
decks.  Each deck is characterized by a certain amount of gain
(delivered on each draw), frequency of loss, and amount of loss.  Two
decks (A and B) yield consistently high rewards, but also high,
probabilistic penalties and are both (equally) disadvantageous in the
long run.  The other two decks (C and D) yield consistently smaller
rewards, but also low, probabilistic penalties and are both (equally)
advantageous in the long run.  It is assumed that the ventromedial
prefrontal cortex (VMPFC) is active in the IGT as VMPC patients show
impaired task performance.  Their preference for the decks with
immediate high rewards indicates ``myopia for the future''.

% what is the HDT? developmental trends and relevance?
\citet{Crone2004} designed a developmentally appropriate analogue of
the IGT, the Hungry Donkey Task (HDT), with a similar win and loss
schedule although the absolute amounts were reduced by a factor of 25.
The HDT is a pro-social game inviting the player to assist a hungry
donkey to collect as many apples as possible, by opening one of four
doors.  Again, doors A and B are characterized by a high constant gain
(10 apples), whereas doors C and D deliver a low constant gain (2
apples).  At doors A and C, a loss of 50 apples (A) or 10 apples (C)
is delivered in 50\% of the trials.  For doors B and D, frequency of
loss is only 10\%.  The median loss of doors B and D is 10 and 2,
respectively.  \citet{Crone2004} administered the HDT to children from
four age groups (6-9, 10-12, 13-15, and 18-25 year-olds) and concluded
that children also fail to consider future consequences.

% strategic reanalysis of the HDT: different strategies found by
% finite mixture analysis
A reanalysis of this dataset \cite{Huizenga2007} indicated that
participants might solve the task by sequentially considering the
three dimensions (constant gain, frequency of loss, and amount of
loss) in order to choose a door.  The youngest children in the
dataset mostly seem to focus on the dominant dimension in the task, frequency
of loss, resulting in equal preference for doors B and D. Older
participants seem to use a two-dimensional rule where participants
first focus on the frequency of loss and then consider amount of loss,
resulting in a preference for door D. A third very small subgroup
seems to use an integrative rule where participants combine all three
dimensions in the appropriate way.  Participants using the integrative
rule pick cards from doors C and D, which are advantageous in the long
run.

% problematic aspects of standard analysis
Typical analyses of these data use the last 60 trials in a series of
200 trials.  A silent assumption made in these analyses is
that behavior has stabilized after 140 learning trials; this could
very well be wrong and it is highly likely that there are individual
differences in this learning process.

% proposed analysis here
Here, we analyze IGT learning data from a single participant with the aim to
establish 1) at which trial learning stops, i.e., at which
trial behavior has stabilized, and 2) what strategies participants use
during and at the end of learning.  Models with an increasing number
(from 2 through 5) of latent states were fitted to the time series of
various participants.  The responses were fitted with multinomial
logistic models for the 4 possible choices that participants make in
this task.  There were no covariates on any of the parameters.  The
only constraint imposed on the models' parameters was that
there were designated begin and end states.  This means that the
initial state probabilities were fixed to one for the first state and
to zero for the remaining states.  For the transition matrix this means
that the final state was an absorbing state: the probability of
transitioning out of that state is zero.  This was done to ensure that
there would be a final state in which participants end which provides
the possibility of immediately seeing participants' behavior at the
end of the task.  Models were selected using the Akaike Information
Criterion (AIC; Akaike, 1973).

@@

%We chose to analyze two participants that used different strategies at
%the end of the task as analyzed in the manner proposed by
%\citet{Huizenga2007}.  The first participants' data 
The data were best 
described by a 4-state model.  This model's transition matrix is:
$$
\mat{A} = \begin{pmatrix} 
				0.64 & 0.33 & 0.00 & 0.03 \\
				0.10 & 0.89 & 0.01 & 0.00 \\
				0.05 & 0.00 & 0.95 & 0.00 \\
				0.00 & 0.00 & 0.00 & 1.00
		  \end{pmatrix}
$$
As can be seen from the transition matrix the final state is
absorbing.  States 2 and 3 are also fairly stable with high diagonal
probabilities of 0.89 and 0.95 respectively.  The response parameters
are presented in Table~\ref{tab:igt4} below.

@@

The states have clear interpretations. States 3 and 4 are both
dominated by C responses, and state 4 has a low probability of B
responses. States 1 and 2 are dominated C and D and A and B responses 
respectively. To get a clearer interpretation of these states it is
necessary to consider when they are visited during the learning
process. The Viterbi algorithm (REFERENCE) provides us with the
posterior state sequence, i.e., the sequence of states that the
participant is in at each trial. This sequence is depicted in
Figure~\ref{fig:post4}.

@@

At the start of training there are many switches between states 1 and 
2, indicating that behavior is mostly random choices between all four 
categories, albeit with a slight preference for B and D over A and C
choices. This preference indicates a focus on the frequency of loss
associated with each of the doors as B and D have low frequency of
loss. After this, there is a period of stable state 2 behavior
associated with responses A and B. After that there is stable state 3 
behavior, consisting only of C responses, a short transitional period 
and then state 4 behavior consisting of 94 \% C choices and some B
choices, i.e., almost optimal behavior. The choice proportions and the
model predicted choice proportions are depicted in
Figure~\ref{fig:igt4}.

@@

The Weather Prediction Task \cite<WPT>{Knowlton1994} %(WPT, Knowlton, Squire \& Gluck, 1994) 
is
a probabilistic categorization task in which participants learn to
predict (or categorize) the state of the weather (sun or rain) on the
basis of four ``tarot'' cards (cards with abstract geometrical
patterns).  On a given trial, one, two, or three cues are present.
There are a total of 14 possible cue patterns, and each one is
associated with a particular probability distribution over the states
of the weather.  In order to perform in the task, participants must
predict the weather in accordance with these conditional
probabilities.

\nocite{Knowlton1994}

@@

There are different accounts of probabilistic category learning 
\cite<see e.g.>[for an overview]{Ashby2005}.
According to instance or exemplar learning theories, participants
learn by storing each encountered cue-outcome pair.  When presented
with a probe cue pattern, a response is made by retrieving these 
exemplars are retrieved from memory and weighting them according 
to their similarity to the probe cue pattern. According to associative theories,
participants learn by gradually associating the individual
cues (or cue patterns in configural learning) to the outcomes. Finally, in
rule-learning, participants are taken to extract rules by which to
categorize the different cue patterns.  \cite{Gluck2002} proposed a
number of such rules (or strategies). The rules differ in the way
the cues are combined into a response. A main difference is whether responses are
based on the presence/absence of a single cue, or whether information 
from multiple cues is integrated. \citeauthor{Gluck2002}
formulated all strategies in a deterministic and optimal manner (e.g., 
the multi-cue strategy corresponded to giving the optimal response to
each cue pattern). \citet{Meeter2006}  allowed for probabilistic
responding (a small probability of responding non-optimally) as well
as switches between strategies during learning. However, both strategies 
and response probabilities remained predefined. Alternative non-strategy based 
analyses of the WPT \cite{Lagnado2006,Speekenbrink2008} have estimated 
response strategies through logistic regression, allowing the regression 
coefficients to change smoothly over time.

%associative, rule-based.
Here, we combine the regression and strategy-based approaches, and analyse 
the behavior of a single individual performing the
WPT for 200 trials.  We are particularly interested
in evidence for strategy switching and whether a DMM can recover
strategies in accordance with \citeA{Gluck2002}. We chose to analyse the 
``average'' participant (the participant with performance closest to 
the group average) in a large unpublished dataset. We let each state 
be characterized by a GLM with a Binomial distributed response and 
logistic link function (i.e., a logistic regression model). The regression
coefficients of a model relating the cues to the state of the weather are 
given in column ```validity'' of Table~\ref{tab:WPT}. Note that indentical
coefficients for the model relating the cues to responses would indicate 
probability matching. A maximizing strategy is indicated by more
extreme regression coefficients.

@@

Analyzing the WPT data on an individual level allowed us to precisely estimate idiographic strategies and their progression during learning. Moreover, using DMMs, we can combine an individual and group level analysis, increasing reliability of the individual analyses whilst allowing for substantial individual differences. Previous analyses \cite<e.g.,>{Meeter2006} pre-defined individual strategies, based on participants' self-reports in \citet{Gluck2002}. Interestingly, \citet{Gluck2002} noted that these self-reports often did not correspond to the strategies evident in participants' responses. It is well known that people may find it difficult to verbalize the way in which they integrate multiple sources of information. As such, estimating strategies directly from participants' responses will result in more valid assessment of their response strategies, and provides a direct test of the validity of the strategy set identified by \citeauthor{Gluck2002}.

>>>>

%&LaTeX
\documentclass[a4paper,12pt,man]{apa} % nobf, doc, apa

%\usepackage[english]{babel} % ik krijg anders foute citaties...in Zweeds formaat???

@@

\twoaffiliations{Department of Psychology, University of Amsterdam}
{Department of Cognitive, Perceptual and Brain Sciences, University College London}

@@

% balance scale rules
Jansen \& Maas (2001) applied the catastrophe model to
development of reasoning on the balance scale task (Siegler, 1981).
In the balance scale task participants have to judge which side of a
balance goes down when the number of weights and their distances to
the fulcrum are varied over trials.  Younger children tend to ignore
the distance dimension in this task, and instead focus solely on the
number of weights on each side of the fulcrum.  This strategy for
solving balance scale items is called Rule 1 \citet{Siegler1981}.  Older
children include the distance dimension in determining their response
to balance scale problems; however, they only do so when the weight
dimension does not differ between the sides of the balance, i.e., when
the number of weights is equal on both sides of the balance scale.
This strategy is called Rule 2 \cite{Siegler1981}.

@@

% conditioning and addiction research
Also in animal learning and conditioning, evidence is found for sudden
changes in response behavior.  For example, \citet{Gallistel2004}
found evidence for sudden onset of learning: at the start of the
learning experiment, pigeons did not learn anything and performance
was stable; after a number of trials, learning kicked in and there
were large increases in performance.  \citet{Gallistel2004}
focused on modeling the distribution of onset times: that is, the
trials at which learning suddenly takes off.  A similar interest in
process onset times is found in addiction research.  For example,
\citet{Sher2004} study the age at which children start using alcohol
and how this related to eventual outcomes in terms of addiction.

% discrimination and categorization learning
Sudden transitions in learning are also observed in simple
discrimination learning paradigms in which participants learn to
discriminate a number of stimuli based on a single dimension such as
form or color.  \citet{Raijmakers2001} found evidence for two
different strategies applied by children when faced with such a
learning task.  \citet{Schmittmann2006} reanalyzed their data using
hidden Markov models to show that both strategies are characterized by
sudden transitions in the learning process.

% latent Markov versus time series analyses/outline
In above mentioned applications, the data consist mostly of a few
repeated measurements administered to large groups of participants.
The focus in the current chapter is rather on data that consist of
many repeated measurements, or time series, observed in only a few
participants or even just a single participant.  For example,
\citet{Visser2009} analyzed data from a single participant in an
experimental task that manipulates the trade-off between speed and
accuracy.  The data consisted of three time series with each around
150 repeated measurements of both reaction time and accuracy.  Below
we provide examples of analyzing time series from single participants
from two experiments; one from the Iowa Gambling Task and one from the
Weather Prediction Task.  The interest in these tasks is to show that
participants develop different strategies over time in responding to
the stimuli and that the transition from one strategy to the next is a
discrete event.  Before providing these illustrations, we give a
formalization of dependent mixture models and a brief overview of the
\pkg{depmixS4} package that was developed to specify and fit such
models.

@@

% history of Markov models/LMM applications
Markov models have been used in the social sciences; for example, in
analyzing language learning \cite{Miller1952,Miller1963} and in the
analysis of paired associate learning \cite{Wickens1982}.  In these
models, the focus is on survey type data: a few repeated measurements
taken in a large sample.  \citeA{Langeheine1990} discuss latent Markov
models and their use in sociology and political science (see also
McCutcheon, 1987).  Latent Markov models have been used in studying
development of math skills \citep{Collins1992}, and in medical
applications \citep{Reboussin1998}; \citet{Kaplan2008} provides an
overview of such models, also called stage-sequential models in the
developmental psychology literature.

@@

% hidden Markov models
Hidden Markov models (HMMs) tend to be used in the analysis of long
univariate (individual) timeseries.  For example, HMMs are the model
of choice in speech recognition applications \cite{Rabiner1989}.  In
biology, HMMs are used to analyze DNA sequences \cite{Krogh1998} and
in econometric science, to analyze changes in stock market prices and
commodities \cite{Kim1994}.

% depmix model
The dependent mixture model (DMM) that we propose here spans the range
from latent Markov models for few repeated measurements with many
participants to hidden Markov models for individual time series.  In
addition, the dependent mixture model includes multivariate responses
and covariates.  The dependent mixture model for a time series $O_{t}$
consists of the following elements:
\begin{enumerate}
	\item $S$ is a collection of discrete states
	\item $S_{t} = \mat{A}S_{t-1}+\xi_{t}, \mat{A}$, a transition matrix
	\item $O_{t} = \mat{B}(S_{t}) + \zeta_{t}, \mat{B}$,  an observation density.
\end{enumerate}
Here $\xi_{t}$ and $\zeta_{t}$ are independent error processes 
\cite{Elliott1995}.

% state space
The state space, which is a set of discrete states, captures the
different states that the learning or developmental process under
consideration can be in.  In the balance scale example mentioned
above, children are applying one of two possible rules in responding
to the items.  The states are characterized by their corresponding
observation densities.  Using a particular rule in the balance scale
task leads to correct answers on some items and incorrect answers on
others.  A different rule may lead to correct answers on some items
and to guessing behavior on other items.  As another example, in
analyzing categorization learning data, in which participants learn to
categorize a set of objects, a typical initial state is that
participants are guessing because at the start of the task they have
no knowledge of which features are important in categorization.
Hence, the states in the state space represent knowledge states of the
participants, such that different knowledge states lead to different
observed behaviors or responses.

% transitions/matrix
Transition matrix $\mat{A}$ describes the transitions between states
over repeated measurements.  This matrix summarizes the probabilities
of transitioning from one state to another, which represents learning
or development.  The transition model contains the Markov assumption:
$$Pr(S_{t}|S_{t-1}, \ldots, S_{1}) = Pr(S_{t}|S_{t-1}),$$
where $Pr$ is the probability.  The Markov assumption means that the
current state (at time $t$) only depends on the previous state
$S_{t-1}$, and not on earlier states.

% observation densities
The observation densities $\mat{B}$ form the measurement part of the
model; these describe the distributions of the observations
conditional on the current state. Hence, these distributions
characterize the state, and in our examples, these characterize the
rule that a participant is using at a given measurement occasion.

% log-likelihood: kan dit helemaal weg???
The log-likelihood of DMMs is usually computed by the
so-called forward-backward algorithm \citep{Baum1966,Rabiner1989}, or
rather by the forward part of this algorithm.  \citet{Lystig2002}
changed the forward algorithm in such a way as to allow computing the
gradients of the log-likelihood at the same time.  They start by
rewriting the likelihood as follows (for ease of exposition the
dependence on the model parameters is dropped here):
\begin{equation}
	L_{T} = Pr(\vc{O}_{1}, \ldots, \vc{O}_{T}) = \prod_{t=1}^{T}
Pr(\vc{O}_{t}|\vc{O}_{1},
	\ldots, \vc{O}_{t-1}),
	\label{condLike}
\end{equation}
where $Pr(\vc{O}_{1}|\vc{O}_{0}):=Pr(\vc{O}_{1})$. Note that for a
simple, i.e.\ observed, Markov chain these probabilities reduce to
$Pr(\vc{O}_{t}|\vc{O}_{1},\ldots,
\vc{O}_{t-1})=Pr(\vc{O}_{t}|\vc{O}_{t-1})$.
The log-likelihood can now be expressed as:
\begin{equation}
	l_{T} = \sum_{t=1}^{T} \log[Pr(\vc{O}_{t}|\vc{O}_{1}, \ldots,
\vc{O}_{t-1})].
	\label{eq:condLogl}
\end{equation}

@@

	\item to be able to fit transition models with covariates, i.e.,
	such that the transition probabilities depend on a time-dependent 
	covariate.

@@

% response models
Response models in \pkg{depmixS4} interface with the \code{glm} 
functions available in R and with the \pkg{nnet}-package for the 
multinomial logistic models. Normally distributed data can hence be 
modelled with direct effects included as well. For example, normal 
data are modelled as:
\begin{equation}
	O_{t}|S_{t} = \mu + \mathbf{\beta}^{`}\vc{x} + \epsilon_{t},
\end{equation}
where $\mu$ is the (state-dependent) mean; $\vc{x}$ is a vector of
covariates or predictors; $\mathbf{\beta}$ is the vector of
(state-dependent) regression coefficients; $\epsilon_{t}$ is a
normally distributed error term with (state-dependent) standard
deviation $\sigma$.

@@

% referentie: nnet

@@

@@

% what is the IGT?
The Iowa gambling task (IGT) is an experimental paradigm designed to
mimic real-life decision-making situations \cite{Bechara1994}, in the
way that it factors uncertainty, reward and punishment
\cite{Dunn2006}.  The task requires the selection of cards from four
decks.  Each deck is characterized by a certain amount of gain
(delivered on each draw), frequency of loss, and amount of loss.  Two
decks (A and B) yield consistently high rewards, but also high,
probabilistic penalties and are both (equally) disadvantageous in the
long run.  The other two decks (C and D) yield consistently smaller
rewards, but also low, probabilistic penalties and are both (equally)
advantageous in the long run.  It is assumed that the ventromedial
prefrontal cortex (VMPFC) is active in the IGT as VMPC patients show
impaired task performance.  Their preference for the decks with
immediate high rewards indicates ``myopia for the future''
\cite{Bechara1994}.

% what is the HDT? developmental trends and relevance?
\citet{Crone2004} designed a developmentally appropriate analogue of
the IGT, the Hungry Donkey Task (HDT), with a similar win and loss
schedule although the absolute amounts were reduced by a factor of 25.
The HDT is a pro-social game inviting the player to assist a hungry
donkey to collect as many apples as possible, by opening one of four
doors.  Again, doors A and B are characterized by a high constant gain
(10 apples), whereas doors C and D deliver a low constant gain (2
apples).  At doors A and C, a loss of 10 apples (A) or 2 apples (C)
is delivered in 50\% of the trials.  For doors B and D, frequency of
loss is only 10\%.  The median loss of doors B and D is 50 and 10,
respectively.  \citet{Crone2004} administered the HDT to children from
four age groups (6-9, 10-12, 13-15, and 18-25 year-olds) and concluded
that children have difficulty to take future consequences into account.

% strategic reanalysis of the HDT: different strategies found by
% finite mixture analysis
A reanalysis of this dataset \cite{Huizenga2007b} indicated that
participants might solve the task by sequentially considering the
three dimensions (constant gain, frequency of loss, and amount of
loss) in order to choose a door.  Most of the youngest children seem
to focus frequency of loss, resulting in equal preference for doors B
and D. It's assumed that frequency of loss is the dominant dimension
in the task.  Older participants seem to use a two-dimensional rule
where participants first focus on the frequency of loss and then
consider amount of loss, resulting in a preference for door D. A third
very small subgroup seems to use an integrative rule where
participants combine all three dimensions in the appropriate way.
Participants using the integrative rule pick cards from doors C and D,
which are advantageous in the long run.

% Check ref Huizenga 2007b

% problematic aspects of standard analysis
\citet{Huizenga2007b} used the last 60 trials in a series of 200
trials, assuming that behavior has stabilized after 140 trials of
learning; this could very well be wrong and it is highly likely that
there are individual differences in this learning process.  Hence,
here we analyze IGT learning data from a single participant with the
aim to establish 1) at which trial learning stops, i.e., at which
trial behavior has stabilized, and 2) what strategies participants use
during and at the end of learning.  Models with an increasing number
(from 2 through 5) of latent states were fitted to the time series of
various participants.  The responses were fitted with multinomial
logistic models for the four possible choices that participants make in
this task.  There were no covariates on any of the parameters.  The
only constraint imposed on the models' parameters was that there were
designated begin and end states.  This means that the initial state
probabilities were fixed to one for the first state and to zero for
the remaining states.  For the transition matrix this means that the
final state was an absorbing state: the probability of transitioning
out of that state is zero.  This was done to ensure that there would
be a final state in which participants end which provides the
possibility of immediately seeing participants' behavior at the end of
the task.  Models were selected using the Akaike Information Criterion
(AIC; Akaike, 1973).

@@

The data were best described by a 4-state model.  This model's
transition matrix is:
$$
\mat{A} = \begin{pmatrix} 
				0.64 & 0.33 & 0.00 & 0.03 \\
				0.10 & 0.89 & 0.01 & 0.00 \\
				0.05 & 0.00 & 0.95 & 0.00 \\
				0.00 & 0.00 & 0.00 & 1.00
		  \end{pmatrix}
$$
As can be seen from the transition matrix the final state is
absorbing.  States 2 and 3 are also fairly stable with high diagonal
probabilities of 0.89 and 0.95 respectively.  The response parameters
are presented in Table~\ref{tab:igt4} below.

% Table refs checken

@@

The states have clear interpretations.  States 3 and 4 are both
dominated by C responses, and state 4 has a low probability of B
responses.  States 1 and 2 are dominated by C and D, and by A and B
responses, respectively.  To get a clearer interpretation of these
states it is necessary to consider when they are visited during the
learning process.  The Viterbi algorithm \citep{Rabiner1989} provides
us with the posterior state sequence, i.e., the sequence of states
that the participant is in at each trial.  This sequence is depicted
in Figure~\ref{fig:post4}.

% figure refs checken

@@

At the start of training there are many switches between states 1 and 
2, indicating that behavior is mostly random choices between all four 
categories, albeit with a slight preference for B and D over A and C
choices. This preference indicates a focus on the frequency of loss
associated with each of the doors as B and D both have low frequency of
loss. After this, there is a period of stable state 2 behavior
associated with responses A and B. After that there is stable state 3 
behavior, consisting only of C responses, a short transitional period 
and then state 4 behavior consisting of 94 \% C choices and some B
choices, i.e., almost optimal behavior. The choice proportions and the
model predicted choice proportions are depicted in
Figure~\ref{fig:igt4}.

@@

The Weather Prediction Task \cite<WPT>{Knowlton1994} %(WPT, Knowlton, Squire \& Gluck, 1994) 
is a probabilistic categorization task in which participants learn to
predict (or categorize) the state of the weather (sun or rain) on the
basis of four ``tarot'' cards (cards with abstract geometrical
patterns).  On a given trial, one, two, or three cues are present.
There are a total of 14 possible cue patterns, and each cue pattern is
associated with a particular probability distribution over the states
of the weather.  In order to perform well in the task, participants
must predict the weather in accordance with these conditional
probabilities.

@@

There are different accounts of probabilistic category learning
\citep<see e.g.>[for an overview]{Ashby2005}.  According to instance or
exemplar learning theories, participants learn by storing each
encountered cue-outcome pair.  When presented with a probe cue
pattern, a response is made by retrieving these exemplars are
retrieved from memory and weighting them according to their similarity
to the probe cue pattern.  According to associative theories,
participants learn by gradually associating the individual cues (or
cue patterns in configural learning) to the outcomes.  Finally, in
rule-learning, participants are taken to extract rules by which to
categorize the different cue patterns.  \citet{Gluck2002} proposed a
number of such rules (or strategies).  The rules differ in the way the
cues are combined into a response.  A main difference is whether
responses are based on the presence/absence of a single cue, or
whether information from multiple cues is integrated.
\citet{Gluck2002} formulated all strategies in a deterministic
and optimal manner (e.g., the multi-cue strategy corresponded to
giving the optimal response to each cue pattern).  \citet{Meeter2006}
allowed for probabilistic responding (a small probability of
responding non-optimally) as well as switches between strategies
during learning.  However, both strategies and response probabilities
remained predefined.  Alternative non-strategy based analyses of the
WPT \cite{Lagnado2006,Speekenbrink2008} have estimated response
distributions through logistic regression, allowing the regression
coefficients to change smoothly over time.

%associative, rule-based.
Here, we combine the regression and strategy-based approaches, and
analyse the behavior of a single individual performing the WPT for 200
trials.  We are particularly interested in evidence for strategy
switching and whether a DMM can recover strategies in accordance with
\citet{Gluck2002}.  We chose to analyse the ``average'' participant
(the participant with performance closest to the group average) in a
large unpublished dataset.  We let each state be characterized by a
GLM with a Binomial distributed response and logistic link function
(i.e., a logistic regression model).  The regression coefficients of a
model relating the cues to the state of the weather are given in
column ```validity'' of Table~\ref{tab:WPT}.  Note that indentical
coefficients for the model relating the cues to responses would
indicate probability matching.  A maximizing strategy is indicated by
more extreme regression coefficients.

@@

% IGT
confirmation of existence of different strategies

dynamics of learning now known

different strategies/rules observed during learning (eg AB and 
guessing strategies)

% WPT
Analyzing the WPT data on an individual level allowed us to precisely
estimate idiographic strategies and their progression during learning.
Moreover, using DMMs, we can combine an individual and group level
analysis, increasing reliability of the individual analyses whilst
allowing for substantial individual differences.  Previous analyses
\cite<e.g.,>{Meeter2006} pre-defined individual strategies, based on
participants' self-reports in \citet{Gluck2002}.  Interestingly,
\citet{Gluck2002} noted that these self-reports often did not
correspond to the strategies evident in participants' responses.  It
is well known that people may find it difficult to verbalize the way
in which they integrate multiple sources of information.  As such,
estimating strategies directly from participants' responses will
result in more valid assessment of their response strategies, and
provides a direct test of the validity of the strategy set identified
by \citet{Gluck2002}.
